---
title: "WoLKE-Studie"
author: "Benedikt Beuttler"
date: "`r Sys.Date()`"
lang: en
page-layout: full
document-css: false
link-citations: true
number-sections: true
toc: true
csl: https://www.zotero.org/styles/apa
bibliography: library.bib
editor: source
execute:
  cache: false
format:
  html:
    standalone: true
    embed-resources: true
    section-divs: true
    html-math-method: mathjax
    wrap: none
    default-image-extension: png
    date-format: long  
    code-fold: true
    code-tools: true
  docx: 
    toc: true
    toc-depth: 2
    number-sections: true
    fig-caption: true
    fig-format: pdf
    default-image-extension: pdf
    date-format: long
    keep-md: true
    echo: false
    code-fold: true
prefer-html: true
---

# Merging 2025 and 2026 data

This document contains the analysis scripts and report for the manuscript "A Course on AI for Pre-Service STEM and Language Teachers: A Quasi-Experimental Evaluation of Learning Outcomes Using Intelligent-TPACK and AI Literacy Instruments ". The R scripts, data are hosted at <OSF>.

```{r Information for first time user to get started with renv}
#| output: false
if (FALSE) {
  install.packages("renv") #1
  renv::restore() #2
}
```

# Preamble

## Prepare the Environment

## Load Libraries

```{r load-libraries-and-functions, cache=F}
#| output: false

# install pacman
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}
library(pacman)

## Pfad/Utilities
p_load(this.path)

## Wrangling
# use pacman to load packages
p_load(dplyr, tidyr, stringr, forcats, purrr, magrittr)

## Ähnlichkeit/Text
p_load(stringdist, psych)

## Modellierung
p_load(lavaan, lme4, lmerTest, emmeans, estimatr)

## Tabellen/Reporting
# install.packages("remotes")

# remotes::install_version(
#   "modelsummary",
#   version = "2.5.0",
#   repos = "https://cran.r-project.org"
# )

p_load(modelsummary)
p_load(kableExtra, arsenal)

## Visualisierung/Themes
p_load(ggplot2, ggsci, jtools)

p_load(mnormt)

source(file = "functions.r") # all outsourced funtions
```

```{r prepare-output}

is_word <- !(knitr::is_html_output() | knitr::is_latex_output())

if (is_word) {
  p_load(flextable)
}

```

## Load Pre- and Posttest and Reshape Data

```{r analysis-setup}
#define vars
covars <- c(
  "time",
  "gender",
  "semester",
  "location",
  "intervention",
  "age",
  "students",
  "group"
)
controls <- c("gender", "semester", "location")
random_terms_lmer <- c("(1 | code)")
groups <- c("intervention")
time <- "time"
outcomes <- c(
  "AILIT",
  "AILIT26",
  "AILITwolke",
  "AILITglat",
  "TK",
  "TPK",
  "TCK",
  "TPACKce",
  "TPACKgenai",
  "Ethics",
  "CEA",
  "CEA_understanding",
  "CEA_ped",
  "PU"
)
outcomes_ailit <- c("AILIT", "AILIT26", "AILITwolke", "AILITglat")
outcomes_tpack <- c("TK", "TPK", "TCK", "TPACKce", "TPACKgenai", "Ethics")
q_column_names <- c("AI", "GL", "TP", "RE02", "ED", "NP")
vars_num <- c("age", "semester") # numeric outcomes
vars_cat <- c("gender", "students") # categorical outcomes`
```

```{r load and prepare data}

# load pre-prepared data
# 25
load("data/25/25-data_all.Rdata") # nicht nach pre und posttest gefiltert
load("data/25/25-data.Rdata") # probanden mit pre UND posttest
# 26
load("data/26/26-data_all.Rdata") # nicht nach pre und posttest gefiltert
load("data/26/26-data.Rdata") # probanden mit pre UND posttest
# load all Rdata from folder /data/dropouts
files <- list.files(
  "data/26/dropouts",
  pattern = "\\.Rdata$",
  full.names = TRUE
)
invisible(lapply(files, load, envir = .GlobalEnv))

data_merged <- bind_rows(
  data_25 %>%
    select(
      code,
      starts_with(q_column_names, ignore.case = FALSE),
      all_of(covars)
    ) %>%
    mutate(
      wave = "sose25",
      across(c(starts_with(q_column_names), semester, age), as.integer)
    ),

  data_26 %>%
    select(
      code,
      starts_with(q_column_names, ignore.case = FALSE),
      all_of(covars)
    ) %>%
    mutate(
      wave = "wise26",
      across(c(starts_with(q_column_names), semester, age), as.integer)
    )
)

data_all_merged <- bind_rows(
  data_all_25 %>%
    select(
      code,
      starts_with(q_column_names),
      all_of(covars)
    ) %>%
    mutate(
      wave = "sose25",
      across(c(starts_with(q_column_names), semester, age), as.integer)
    ),

  data_all_26 %>%
    select(
      code,
      starts_with(q_column_names),
      all_of(covars)
    ) %>%
    mutate(
      wave = "wise26",
      across(c(starts_with(q_column_names), semester, age), as.integer)
    )
)

data_merged_pre <- data_merged %>% filter(time == "pre")
data_all_merged_pre <- data_all_merged %>% filter(time == "pre")

```

```{r Prepare scales}

### grouping scales
## TPACK
# all TPACK items used
list_items_i_tpack <- list(
  "TK" = grep("^TP01_", names(data_merged), value = T),
  "TPK" = grep("^TP02_", names(data_merged), value = T),
  "TCK" = grep("^TP03_", names(data_merged), value = T),
  "TPACK_ce" = grep("^TP04_(0[1-8]|12|15)$", names(data_merged), value = T), # Sub-Subskala 1: Celik Enhanced Items
  "TPACK_genai" = grep(
    "^TP04_(09|10|11|13|14)$",
    names(data_merged),
    value = T
  ), # Sub-Subskala 2: GenAI Items
  "Ethics" = grep("^TP05_0[1-6]$", names(data_merged), value = T) #im pretest nur items 1-6 vorhanden 7,8 kamen im post dazu
)

## Perceived Usefulnes
list_items_pu <- list(
  "Pu" = grep("^ED03_", names(data_merged), value = T)
)

# ein paar ethik items müssen noch rausgenommen werden, die nur im posttest erhoben wurden. hier nur check
# data_subset1 <- data %>% select(all_of(unlist(list_items_i_tpack)))
# View(data_subset1)

## Critical Evaluation of AI
list_items_cea <- list(
  "CEA" = grep("^RE02_(0[1-9])$", names(data_merged), value = T),
  "CEA_understanding" = grep("^RE02_(0[1-4])$", names(data_merged), value = T),
  "CEA_ped" = grep("^RE02_(0[5-9])$", names(data_merged), value = T)
)
# data_subset3 <- data %>% select(all_of(unlist(list_items_cea)))
# sum(data_subset3 < 0, na.rm = TRUE)

## AI Literacy
ailit_wolke_items <- c(
  "AI36_04",
  "AI36_10",
  "AI36_11",
  "AI36_13",
  "AI36_15",
  "AI36_16",
  "AI36_17",
  "AI36_19",
  "AI36_24",
  "AI36_25",
  "AI36_27"
)
ailit_glat_items <- c(
  "GL21_01",
  "GL21_02",
  "GL21_03",
  "GL21_04",
  "GL21_05",
  "GL21_06",
  "GL21_07",
  "GL21_08",
  "GL21_09",
  "GL21_10",
  "GL21_11",
  "GL21_12",
  "GL21_13",
  "GL21_14",
  "GL21_15",
  "GL21_16",
  "GL21_17",
  "GL21_18",
  "GL21_19"
)

list_items_ailit <- list(
  "ailit" = grep(
    paste0(
      "^(",
      paste(c(ailit_wolke_items, ailit_glat_items), collapse = "|"),
      ")$"
    ),
    names(data_merged),
    value = TRUE
  ),
  "ailit_wolke" = grep(
    paste0("^(", paste(ailit_wolke_items, collapse = "|"), ")$"),
    names(data_merged),
    value = TRUE
  ),
  "ailit_glat" = grep(
    paste0("^(", paste(ailit_glat_items, collapse = "|"), ")$"),
    names(data_merged),
    value = TRUE
  )
)

# create new columns with the means of the lists
data_merged <- data_merged %>%
  mutate(
    AILIT = rowSums(cbind(AI35_02, GL20_01), na.rm = TRUE), # default sumscore overall AI Lit
    AILIT26 = if_else(
      wave == "wise26",
      rowSums(cbind(AI35_02, GL20_01), na.rm = TRUE),
      NA_real_
    ),
    AILITwolke = rowSums(
      select(., all_of(list_items_ailit$ailit_wolke)),
      na.rm = TRUE
    ),
    AILITglat = GL20_01,
    TK = rowMeans(select(., all_of(list_items_i_tpack$TK)), na.rm = TRUE),
    TPK = rowMeans(select(., all_of(list_items_i_tpack$TPK)), na.rm = TRUE),
    TCK = rowMeans(select(., all_of(list_items_i_tpack$TCK)), na.rm = TRUE),
    TPACKce = rowMeans(
      select(., all_of(list_items_i_tpack$TPACK_ce)),
      na.rm = TRUE
    ),
    TPACKgenai = rowMeans(
      select(., all_of(list_items_i_tpack$TPACK_genai)),
      na.rm = TRUE
    ),
    Ethics = rowMeans(
      select(., all_of(list_items_i_tpack$Ethics)),
      na.rm = TRUE
    ),
    TPACK_total = rowMeans(
      select(., all_of(unlist(list_items_i_tpack))),
      na.rm = TRUE
    ),
    CEA = rowMeans(select(., all_of(list_items_cea$CEA)), na.rm = TRUE),
    CEA_understanding = rowMeans(
      select(., all_of(list_items_cea$CEA_understanding)),
      na.rm = TRUE
    ),
    CEA_ped = rowMeans(select(., all_of(list_items_cea$CEA_ped)), na.rm = TRUE),
    PU = rowMeans(select(., all_of(list_items_pu$Pu)), na.rm = TRUE),
    PU1 = ED03_01,
    PU2 = ED03_02,
    PU3 = ED03_03,
    PU4 = ED03_04
  )

# check missing values in TPACK and PU scales (NA means, not answered (-1 diff. answer, -9 not answered)) -> this comes from ethic items
data_merged %>%
  select(all_of(c(list_items_i_tpack) %>% unlist)) %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  sum # 0

```

```{r reshape data}

data_merged_long <- data_merged %>%
  pivot_longer(
    cols = all_of(outcomes),
    names_to = "outcome",
    values_to = "score"
  ) %>%
  mutate(
    outcome = factor(outcome, levels = outcomes)
  )

```

```{r descriptive-plots}
# TPACK
p_tpack <- plot_outcome(
  data_merged_long %>% filter(str_detect(outcome, "^(T)")),
  scales_facet = "fixed"
) +
  ggtitle("Intelligent-TPACK")

ggsave(
  p_tpack + facet_wrap(~outcome, ncol = 3),
  file = "figures/plot-tpack.pdf",
  width = 8,
  height = 10,
  device = cairo_pdf
)

# AILIT
p_ailit <- plot_outcome(
  data_merged_long %>% filter(str_detect(outcome, "^(AILIT)"))
) +
  ggtitle("AI Literacy")

# Critical Evaluation of AI
p_cea <- plot_outcome(
  data_merged_long %>% filter(str_detect(outcome, "^(CEA)"))
) +
  ggtitle("Critical Evaluation of AI (CEA)")

p_pu <- plot_outcome(
  data_merged_long %>% filter(str_detect(outcome, "^(PU)")),
  scales_facet = "fixed"
) +
  ggtitle("Perceived Usefulness")

```

# Methods

## Participants

```{r participants, cache=F}
p_load(dplyr, tidyr) # needs to be called for rendering to work

data_participants <- data_merged %>%
  filter(time == "pre") %>%
  group_by(intervention, location, gender) %>%
  summarise(
    n = n(),
    n_complete = sum(complete.cases(.)),
    age_min = min(age, na.rm = T),
    age_max = max(age, na.rm = T),
    age_mean = mean(age, na.rm = T) %>% round(2),
    age_sd = sd(age, na.rm = T) %>% round(2),
    semester_min = min(semester, na.rm = T),
    semester_max = max(semester, na.rm = T),
    semester_mean = mean(semester, na.rm = T) %>% round(2),
    semester_sd = sd(semester, na.rm = T) %>% round(2),
    .groups = "drop"
  )

data_participants_25 <- data_merged %>%
  filter(time == "pre", wave == "sose25") %>%
  group_by(intervention, location, gender) %>%
  summarise(
    n = n(),
    n_complete = sum(complete.cases(.)),
    age_min = min(age, na.rm = T),
    age_max = max(age, na.rm = T),
    age_mean = mean(age, na.rm = T) %>% round(2),
    age_sd = sd(age, na.rm = T) %>% round(2),
    semester_min = min(semester, na.rm = T),
    semester_max = max(semester, na.rm = T),
    semester_mean = mean(semester, na.rm = T) %>% round(2),
    semester_sd = sd(semester, na.rm = T) %>% round(2),
    .groups = "drop"
  )

data_participants_26 <- data_merged %>%
  filter(time == "pre", wave == "wise26") %>%
  group_by(intervention, location, gender) %>%
  summarise(
    n = n(),
    n_complete = sum(complete.cases(.)),
    age_min = min(age, na.rm = T),
    age_max = max(age, na.rm = T),
    age_mean = mean(age, na.rm = T) %>% round(2),
    age_sd = sd(age, na.rm = T) %>% round(2),
    semester_min = min(semester, na.rm = T),
    semester_max = max(semester, na.rm = T),
    semester_mean = mean(semester, na.rm = T) %>% round(2),
    semester_sd = sd(semester, na.rm = T) %>% round(2),
    .groups = "drop"
  )


```

The study involved pre-service teachers from two universities of education in Baden-Württemberg, Germany. Participants were divided into a **treatment group** consisting of two intervention groups (STEM and language course), and a **control group**.
Students in the treatment groups self-selected the courses "AI in STEM/Language Education" as part of their compulsory elective studies. The control group was recruited from other STEM and language courses at the Universities during the same semester. (genauer drauf eingehen, dass sprachgruppe nur in lubu kontrolle hatte?)

In total, $N=`r data_all %>% filter(time=="pre") %>% nrow()`$ students originally participated in the study ($N_c=`r data_all %>% filter(time=="pre", intervention=="control") %>%nrow`$ in the control group, $N_s= `r data_all %>% filter(time=="pre", intervention=="stem") %>%nrow `$ in the STEM course, and $N_l=`r data_all %>% filter(time=="pre", intervention=="language") %>%nrow `$ in the language course), of which $n=`r data_participants$n %>% sum`$ completed both pre- and the posttest and were included in the analyses ($n_c = `r data_participants%>%filter(intervention=="control") %>% select(n)%>%sum`$, $n_s = `r data_participants%>%filter(intervention=="stem") %>% select(n)%>%sum`$, $n_l = `r data_participants%>%filter(intervention=="language") %>% select(n) %>% sum`$).

Unfortunately, we observed a high drop-out rate between the pre- and posttest. We attribute this to several factors: First, the posttests were conducted in class close  (one to two weeks) to the examination period, resulting in students not attending the in-person posttest sessions. For the control group, some lecturers executed the posttest during their teaching, while others did not. We suspect a high drop-out at the posttest for the control group due to both factors: students not attending the sessions close to the examination week and not conducting the tests at home or in their free study time.

Descriptive statistics of students' demographics and pretest scores are presented in [@tbl-participants]. The final sample size included in the analyses was $n=`r nrow(data_pre)`$, aged between $`r data_pre$age%>%min`$ and $`r data_pre$age %>% max`$ years (true facts?) ($M=`r data_pre$age%>%mean %>% round(2)`$, $SD=`r data_pre$age%>%sd%>%round(2)`$), of which $`r data_pre%>%filter(gender=="Female")%>%nrow`$ identified as female and $`r data_pre%>%filter(gender=="Male")%>%nrow`$ as male.
The students were enrolled in semester $`r data_pre$semester%>%min`$ to $`r data_pre$semester%>%max`$ ($M=`r data_pre$semester%>%mean%>%round(2)`$, $SD=`r data_pre$semester%>%sd%>%round(2)`$). Of these, $`r data_pre%>%filter(students=="bachelor")%>%nrow`$ students were at the Bachelor's level and $`r data_pre%>%filter(students=="master")%>%nrow`$ at the Master's level.

```{r table-participants, result = "asis"}
#| label: tbl-participants
#| tbl-cap: "Demographic Data of the Three Experimental Groups."
# render as table
format <- ifelse(
  knitr::is_html_output(),
  "markdown",
  ifelse(knitr::is_latex_output(), "latex", "markdown")
)

var_names <- list(
  gender = "Gender",
  age = "Age",
  semester = "Semester Number",
  students = "Study Level",
  isp = "Completed Integrated Semester Internship",
  subject1 = "Subject 1",
  subject2 = "Subject 2",
  studies = "Study Program",
  location = "University",
  TK = "Intelligent TK",
  TCK = "Intelligent TCK",
  TPK = "Intelligent TPK",
  TPACK_total = "Intelligent TPACK",
  AILIT = "AI Literacy",
  Ethics = "AI Ethics",
  CEA = "Critical Evaluation of AI",
  PU = "Perceived Usefulness"
)
# total
kable(
  summary(
    tableby(
      intervention ~
        location +
        age +
        gender +
        semester +
        students,
      data_merged %>%
        filter(time == "pre") %>%
        mutate(
          students = factor(students, labels = c("Bachelor", "Master"))
        ) %>%
        droplevels(),
      cat.test = "chisq",
      chisq.correct = T,
      control = tableby.control(test = T, total = T, digits = 2)
    ),
    labelTranslations = var_names,
    pfootnote = T,
    text = F
  ),
  escape = T,
  format = format,
  booktabs = T
)
# 25
kable(
  summary(
    tableby(
      intervention ~
        location +
        age +
        gender +
        semester +
        students +
        TK +
        TCK +
        TPK +
        TPACK_total +
        Ethics +
        AILIT +
        PU,
      data_merged %>%
        filter(time == "pre", wave == "sose25") %>%
        mutate(
          students = factor(students, labels = c("Bachelor", "Master"))
        ) %>%
        droplevels(),
      cat.test = "chisq",
      chisq.correct = T,
      control = tableby.control(test = T, total = T, digits = 2)
    ),
    labelTranslations = var_names,
    pfootnote = T,
    text = F
  ),
  escape = T,
  format = format,
  booktabs = T
)
# 26
kable(
  summary(
    tableby(
      intervention ~
        location +
        age +
        gender +
        semester +
        students +
        TK +
        TCK +
        TPK +
        TPACK_total +
        Ethics +
        AILIT +
        CEA,
      data_merged %>%
        filter(time == "pre", wave == "wise26") %>%
        mutate(
          students = factor(students, labels = c("Bachelor", "Master"))
        ) %>%
        droplevels(),
      cat.test = "chisq",
      chisq.correct = T,
      control = tableby.control(test = T, total = T, digits = 2)
    ),
    labelTranslations = var_names,
    pfootnote = T,
    text = F
  ),
  escape = T,
  format = format,
  booktabs = T
)
```

## Materials

### Translation Process (to be reworked)
The Intelligent-TPACK, originally developed by @celik2023, is widely used to assess teachers’ knowledge for pedagogical and ethical AI use based on the technological, pedagogical, and content knowledge (TPACK) framework.
However, no empirically validated German translations of the Intelligent-TPACK exists. While some studies translate and use subsets of the Intelligent-TPACK [e.g., @mah.gross2024], there is no validated German version of the full instrument. 

Therefore, for the Intelligent-TPACK by @celik2023 as well as for items from the AI ethics subscale from @chiu.ahmad.ea2024, we translated the original items in a participative multi-step process involving not just experts from the fields of computer science and education, but also (pre-service) teachers.
For this, we applied the *Modified Direct Translation* that includes translators and experts discussing the wording until reaching agreement [@behling.law2025]: First, seven researchers and experts in AI ED (additionally including one English teacher and one student teacher of Computer Science and English) translated the items independently. Second, we compared all translations and marked the items that received different translations. In a third step, we performed a group workshop with all translators in which the marked items were discussed and dot voting was applied to decide on the most apporopriate translation. Finally, three items that were still marked as questionable were discussed and further adapted with a native English teacher who teaches at University 2.

The translation and validation of the Intelligent-TPACK is reported in detail in @authorsinprep.


### Data Preparation and Cleaning (needs to be done after data gets merged)

```{r dropouts}
# get participants who only participated in pre or post test
data_dropouts <- data_all_merged %>%
  group_by(code) %>%
  summarise(
    has_pre = any(time == "pre"),
    has_post = any(time == "post"),
    .groups = "drop"
  ) %>%
  summarise(
    n_pre_only = sum(has_pre & !has_post),
    n_post_only = sum(!has_pre & has_post),
    n_both = sum(has_pre & has_post)
  )
```

if the bullet point already exists in the text, take it from there, adjust with following changes:
- very first step was to only allow trials that are completed
- check codes with levenshtein distance 1 and 2 and correct them
- correct multiple post or pre post entries of same codes down to one (students that filled out)

For data cleaning, we applied the following steps: First, we investigated which students, based on their pseudonyms, did not participate in both test sessions. Based on a string distance analysis using the Levenshtein distance [@levenshtein1966], we identified pseudonyms that were likely to belong to the same student but were entered slightly differently (e.g., due to typos or students who moved during the semester, as the address was part of their pseudonym). We then manually checked these pseudonyms and merged them if we believe they belonged to the same student. This resulted in $n=24$ students being merged.
We then removed students who only participated at the pretest ($n=`r data_dropouts$n_pre_only`$) or only at the posttest ($n=`r data_dropouts$n_post_only`$).
Further, some students conducted the post test twice. In this case, we only considered their first response in the analyses and dropped the second.

As for some self-constructed items students could choose "I don't understand the item" as a response option, we recoded these responses as missing values ($n=`r data %>%
  select(all_of(c(list_items_i_tpack, list_items_pu)%>%unlist)) %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>% sum`$ responses to individual items).

For the Intelligent TPACK, AI ethics and perceived usefulness (sub)scales, we calculated the mean score across all items of the respective (sub)scale per student. For AI literacy, we calculated the sum score across all items. We additionally calculated the sum score for the subset of items of the AI literacy   that that directly reflect the topics addressed in the course, see SECTION MATERIALS.


## Statistical Analysis

We analyzed the interventions' efficacy in a multiple linear regression analysis approach. For this, we implemented a one-way ANCOVA for each outcome measure, considering our $3$ factorial design.

::: {.callout-note}
We did not include the university dummy variable for the clustering of students within universities, as we only have control students at University 2 (and thus yields a high correlation with treatment as well).
Kontraste noch nennen?
:::

Additionally, we included the respective pretest score of the outcome measures as a covariate in the analyses.
Finally, we included variables as a covariate if they met the criteria for baseline equivalence in randomized controlled trials by the What Works Clearinghouse [@WWC2022]. That is, we included the variables as covariates if the difference between the groups was statistically significant, or if it exceeded $d>.05$.
We tested for group differences concerning the following variables at baseline (pretest): age, gender, semester, and study level (Bachelor vs. Master). We found for gender and semesters  statistically significant differences between the treamtment groups, see [@tbl-participants]. We thus included the following covariates in the models: gender and semester.

That is, the final model for each outcome measure contained treatment, pretest score, gender, semester, and University as fixed effects, and was specified as follows:
`lm_robust(outcome_post ~ treatment + outcome_pre + gender + semester )`.

Posttest scores were standardized before running the analyses, so that the regression coefficients of the dummy variable indicating the respective treatment groups can directly be interpreted as Cohen’s d [@marsh2009].

To compute heteroscedasticity-consistent (HC) standard errors, we used the *lm_robust* function from the *estimatr* package [@blair2025], specifying the HC3 estimator, as recommended for small samples [@long2000].

The criterion of statistical significance was set at α=.05.

The raw data and *R* scripts detailing the analyses can be accessed in the online supplement at <OSF>.

### Statistical Software
All analyses were conducted in *R* [@R2024]. Data were prepared using *dplyr* [wickham2023:dplyr], *tidyr* [@wickham2024], and *gtools* [@warnes2023]. Tables were created with *arsenal* [@heinzen2021], *vtable* [@huntington2023], *modelsummary* [@arel-bundock2022], and *kableExtra* [@zhu2024]. Figures were produced using *ggplot2* [@wickham2016], *ggsci* [@xiao2024], and *jtools* [@long2022]. The ANCOVA models were estimated using *lm_robust* from the *estimatr* package [@blair2025]. To compute heteroscedasticity-consistent (HC) standard errors, we specified the HC3 estimator, as recommended for small samples [@long2000]. Where applicable, pairwise comparisons were obtained with *emmeans* [@lenth2025].

```{r ancova}
# reshape data to wide format for ancova
data_wide <- data_merged %>%
  pivot_wider(
    id_cols = c(
      code,
      intervention,
      group,
      gender,
      semester,
      students,
      location
    ),
    names_from = time,
    values_from = all_of(outcomes)
  ) %>%
  mutate(
    # scale the posttest score so that the estimates are approximations of Cohen's d
    across(all_of(outcomes %>% paste0("_post")), scale),
    treatment = factor(ifelse(
      intervention == "control",
      "control",
      "intervention"
    ))
  )

contrasts.treatment <- cbind(c(-1 / 2, 1 / 2))
colnames(contrasts.treatment) <- c("-c+t")
contrasts(data_wide$treatment) <- contrasts.treatment

ancovas <- lapply(
  outcomes,
  fit_ancova,
  groups,
  # we remove University as a covariate as we only have control students at University 2
  setdiff(controls, "location"),
  "HC3",
  data_wide
)
names(ancovas) <- outcomes

ancovas_two_groups <- lapply(
  outcomes,
  fit_ancova,
  "treatment",
  setdiff(controls, "location"),
  "HC3",
  data_wide
)
names(ancovas_two_groups) <- outcomes

```

## AI Literacy

```{r fig-ai-lit, echo=FALSE, fig.cap="Descriptive Statistics of Students' AI Literacy Scores Over Time by Treatment Group.", out.width="100%",fig.width=7, fig.height=6.5}
ailit_n_labels <- data_merged_long %>%
  filter(outcome %in% outcomes_ailit) %>%
  filter(!is.na(score)) %>%
  group_by(outcome) %>%
  summarise(n = n_distinct(code), .groups = "drop")

plot_outcome(
  data_merged_long %>%
    filter(outcome %in% outcomes_ailit) %>%
    mutate(
      outcome = factor(
        outcome,
        levels = outcomes_ailit,
        labels = c(
          "AI Literacy - Full Scale incl. 25",
          "AI Literacy - Full Scale 26",
          "AI Literacy - Study-Specific",
          "AI Literacy - GLAT"
        )
      )
    ),
  ylab = "Score"
) +
  geom_text(
    data = ailit_n_labels %>%
      mutate(
        outcome = factor(
          outcome,
          levels = outcomes_ailit,
          labels = c(
            "AI Literacy - Full Scale incl. 25",
            "AI Literacy - Full Scale 26",
            "AI Literacy - Study-Specific",
            "AI Literacy - GLAT"
          )
        )
      ),
    aes(x = 1.5, y = Inf, label = paste0("n = ", n)),
    inherit.aes = FALSE,
    vjust = 1.5
  ) +
  ggtitle("AI Literacy") +
  theme(legend.position = "bottom")

#p_ailit %+% + facet_wrap(~outcome, ncol =2, scales="free")  + labs(y = "Score")  + theme(legend.position = "bottom")

```

```{r tab.layout = "autofit"}
#| label: tbl-ancova-ailit-summary
#| tbl-cap: "Summary of Multiple Regression Models for AI Literacy."

ft_ailit_ancova <- modelsummary_models(
  ancovas[names(ancovas) %in% outcomes_ailit],
  outcomes = outcomes_ailit,
  output = modelsummary_output
) # %>%   kableExtra::scroll_box(width = "100%")

if (is_word) {
  ft_ailit_ancova <- adjust_table_for_word(ft_ailit_ancova, width = .75)
}
ft_ailit_ancova
```


```{r, results="asis"}
if (!is_word) {
  cat("### Pairwise Contrasts Between Groups")
}
```

```{r ailit-ancova-between-contrasts}
# pairwise between contrasts (no multiplicity adjustment)
ailit_ancova_between_contrasts <- lapply(ancovas[1:3], function(m) {
  emmeans(m, ~intervention) %>%
    contrast(interaction = "pairwise", adjust = "none")
})
if (!is_word) {
  print(ailit_ancova_between_contrasts)
}
```


## Intelligent TPACK and AI Ethics

```{r fig-tpack, echo=FALSE, fig.cap="Descriptive Statistics of Students' Intelligent TPACK and AI Ethics Ratings Over Time by Treatment Group.", out.width="100%",fig.width=7, fig.height=6.5}

n_tpack <- data_merged_long %>%
  filter(outcome %in% outcomes_tpack) %>%
  filter(!is.na(score)) %>%
  summarise(n = n_distinct(code)) %>%
  pull(n)

p_tpack +
  facet_wrap(
    ~outcome,
    scales = "fixed",
    ncol = 3,
    labeller = labeller(
      outcome = c(
        TK = "Intelligent TK",
        TCK = "Intelligent TCK",
        TPK = "Intelligent TPK",
        TPACKce = "Intelligent TPACK - Core",
        TPACKgenai = "iTPACK - Generative AI",
        Ethics = "AI Ethics"
      )
    )
  ) +
  labs(
    title = "Intelligent TPACK and AI Ethics",
    subtitle = paste0("N = ", n_tpack)
  ) +
  theme(legend.position = "bottom")
```


::: {.landscape}
```{r tab.layout="autofit"}
#| label: tbl-ancova-tpack-summary
#| tbl-cap: "Summary of Multiple Regression Models for Intelligent TPACK Subscales and AI Ethics"

ft_ancova <- modelsummary_models(
  ancovas[grepl("^(T|Ethics)", names(ancovas))],
  outcomes = outcomes %>% grep("^(T|Ethics)", ., value = TRUE),
  output = modelsummary_output,
) # %>%   kableExtra::scroll_box(width = "100%")

if (is_word) {
  ft_ancova <- adjust_table_for_word(ft_ancova, width = 1)
}
ft_ancova
```

:::

## Critical Evaluation of AI (CEA)
Note für uns: ggf. reduzieren wir CEA auf eine Skala mit 9 Items. Im Piloten haben wir 15 items getestet und EFA hat gezeigt, dass Reduktion auf 2 Skalen möglich wäre. Das haben wir gemacht, und die funktionieren auch, aber ein Item lädt z.B. ziemlich stark auf die andere Skala. Entweder bekommt man das gut begründet, aber den richtigen Mehrwert sehe ich nicht, 2 Skalen aufzustellen. So oder so, es haben alle Skalen gut funktioniert und zeigen den Mehrwert vom Seminar.

```{r fig-cea, echo=FALSE, fig.cap="Descriptive Statistics of Students' Critical Evaluation of AI Over Time by Treatment Group.", out.width="100%",fig.width=7, fig.height=6.5}

n_cea <- data_merged_long %>%
  filter(outcome %in% "CEA") %>%
  filter(!is.na(score)) %>%
  summarise(n = n_distinct(code)) %>%
  pull(n)

p_cea +
  facet_wrap(
    ~outcome,
    scales = "fixed",
    ncol = 2,
    labeller = labeller(
      outcome = c(
        CEA = "Critical Evaluation of AI (CEA) - Full Scale",
        CEA_understanding = "Critical Understanding of AI",
        CEA_ped = "Pedagogical Evaluation of AI Output"
      )
    )
  ) +
  labs(
    title = "Critical Evaluation of AI (CEA)",
    subtitle = paste0("N = ", n_cea)
  ) +
  theme(legend.position = "bottom")
```

::: {.landscape}
```{r tab.layout="autofit"}
#| label: tbl-ancova-CEA-summary
#| tbl-cap: "Summary of Multiple Regression Models for CEA subscale."

ft_cea_ancova <- modelsummary_models(
  ancovas[grepl("^(CEA)", names(ancovas))],
  outcomes = c("CEA", "CEA_understanding", "CEA_ped"),
  output = modelsummary_output,
) # %>%   kableExtra::scroll_box(width = "100%")

if (is_word) {
  ft_cea_ancova <- adjust_table_for_word(ft_cea_ancova, width = .75)
}
ft_cea_ancova
```

:::

``` {r Exploring CEA data}
# cronbach alpha
alphas <- imap(
  list_items_cea,
  ~ psych::alpha(
    dplyr::select(data_merged_pre, dplyr::all_of(.x)),
    warnings = FALSE
  )
)
alphas
# 0.86 alpha, kein item lohnt sich rauszunehmen

data_merged_post <- data_merged %>% filter(time == "post" & CEA != "NaN")
# EFA
ncomp <- 2
# EFA
df_cea_items <- data_merged_post %>%
  select(all_of(list_items_cea$CEA)) %>%
  na.omit()
cea_scale <- list_items_cea %>% unlist(use.names = F)

efa_promax <- df_cea_items %>%
  select(any_of(cea_scale)) %>%
  psych::fa(nfactors = ncomp, rotate = "varimax", fm = "ml", missing = T)

# print(efa_promax, digits = 2, cut = .3, sort = F)
fa.diagram(efa_promax, digits = 2, cut = .30, simple = FALSE)

## Computing EFA using the lavaan package
cea_efa <- lavaan::efa(
  df_cea_items %>%
    select(any_of(cea_scale)),
  nfactors = ncomp,
  rotation = "varimax",
  ordered = F,
  std.lv = TRUE,
  missing = "fiml"
)

summary(cea_efa, nd = 3L, cutoff = 0.3, dot.cutoff = 0.05)

```

Results from efa and cronbachs alpha confirm our exploration and adjustments from pilot data. we have a good scale in CEA.


## Two Groups (Control vs. Intervention)

```{r ancova-two-groups}
#| label: tbl-ancova-two-groups
#| tbl-cap: "Summary of Multiple Linear Regression Models for Two Treatment Groups."

modelsummary_models(
  ancovas_two_groups,
  outcomes = outcomes,
  output = modelsummary_output
)

```


```{r save-env }
# save environment
save.image("data/env-analysis.Rdata")
```